|

### 2021-11-10 submission/manuscript.rmd
first revision of results section

---
title: |
 Linear interaction of treatment with baseline risk was sufficient for predicting
 individualized benefit
abstract: |
 **Objective:** To compare different risk-based methods predicting individualized
 treatment effects in GRIT simulations. **Study Design and Setting:** We simulated
 data using diverse assumptions for a baseline prognostic index of risk (PI) and
 the shape of its interaction with treatment (none, linear or quadratic). In
 each sample we predicted absolute benefit using: models with the PI and a
 constant relative treatment effect; models including an interaction of treatment
 with the PI; stratification in quarters of the PI; nonlinear transformations of
 the PI (restricted cubic splines with 3, 4 and 5 knots); an adaptive approach
 using Akaike’s Information Criterion. We evaluated predictive performance using
 root mean squared error and measures of discrimination and calibration for
 benefit. Starting from a base case scenario (sample size 4,250, treatment odds
 ratio 0.8, AUC of the PI 0.75), we varied the sample size, the treatment effect
 strength, and the Ti's discriminative ability. **Results:** Models including a
 PI by treatment interaction performed better under most simulation settings.
 Flexible models required larger sample sizes and higher AUC of the PI to
 outperform the linear interaction model. The adaptive approach tended to select
 simpler models with smaller sample sizes and/or worse prediction performance.
 **Conclusion:** Under most circumstances, a model with a linear interaction of
 the PI with treatment is the optimal risk-based approach to predicting absolute
 treatment benefit.
authors:
  - name: Alexandros Rekkas
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Peter R. Rijnbeek
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Ewout W. Steyerberg
    department: Department of Biomedical Data Sciences
    affiliation: Leiden University Medical Center
    location: Leiden, The Netherlands
  - name: David van Klaveren
    department: Department of Public Health
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
keywords:
  - treatment effect heterogeneity
  - absolute benefit
  - prediction models
output: 
  rticles::arxiv_article:
    keep_tex: true
  bookdown::word_document2:
    reference_docx: reference.docx
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage[pagewise, left]{lineno}
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{caption}
  - \usepackage{booktabs}
  - \date{}
  - \newcommand\given[1][]{\:#1\vert\:}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

@@

Combining all these settings resulted in a simulation study of 486 scenarios
(exact settings in the supplementary material).

@@

Second, we considered a model which assumes *constant relative treatment effect*
(constant OR). Hence, absolute benefit is predicted from
$\hat{\tau}(\bm{x}) = \text{expit}(PI +\log(\text{OR}))$.

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE}
adaptiveSelections <- readr::read_csv(here::here("data/processed/adaptiveModel.csv"))
adaptiveSelections <- adaptiveSelections %>% 
  group_by(scenarioId) %>% 
  nest() %>% 
  mutate(
    counts = lapply(data, plyr::count),
    n = lapply(data, nrow)
  ) %>%
  select(
    scenarioId, counts, n
  ) %>%
  unnest(c(counts, n)) %>%
  mutate(
    perc = round(freq / n, 2),
    model = case_when(
      selectedAdaptiveModel == "RCS with 3 knots" ~ "rcs_3",
      selectedAdaptiveModel == "RCS with 4 knots" ~ "rcs_4",
      selectedAdaptiveModel == "RCS with 5 knots" ~ "rcs_5",
      TRUE                                        ~ selectedAdaptiveModel
    )
  ) %>%
  ungroup() %>%
  select(scenarioId, model, perc) %>%
  spread(model, perc) %>%
  relocate(c(treatment, risk), .before = rcs_3)

```

The linear interaction model outperformed all RCS methods in terms of
RMSE in scenarios with true constant relative treatment effect (OR =
0.8, N = `r 4250` and AUC = 0.75), strong linear and even strong
quadratic deviations (Figure \ref{fig:rmsebase}). However, with
non-monotonic deviations errors of the linear interaction model
increased substantially, especially in the presence of
treatment-related harms. In these scenarios, RCS-3 proved to be quite
robust outperforming all other methods. The constant treatment effect
approach had overall best performance under true constant treatment
effect settings, but was sensitive to all considered deviations,
resulting in increased error rates. Finally, the adaptive approach had
comparable performance to the best-performing method in each
scenario. However, we observed increased variability in the case of
linear and non-monotonic deviations, especially for moderate or strong
treatment-related harms.

@@

Increasing the sample size to `r 17000` favored the RCS-3 approach the
most, as it achieved lowest or close to lowest RMSE across all
scenarios (Figure \ref{fig:rmsesamplesize}). Especially in cases of
strong quadratic and non-monotonic deviations RCS-3 had lower error
rates compared to the linear interaction approach, regardless of
treatment-related harms strength. The issues with the large error
variability of the adaptive approach improved with larger sample
sizes.

@@

When we increased the AUC of the true prediction model to 0.85, models including
RCS smoothing had the lowest RMSE in the presence of strong quadratic deviations
from the base case of true constant relative treatment effects (Figure
\ref{fig:rmseauc}; Panel D). However, with milder deviations, the linear
interaction model had the lowest RMSE with the RCS smoothing methods (3 knots)
being a close second (Figure \ref{fig:rmseauc}; Panels B and C). Increasing the
number of knots of RCS smoothing resulted in increased RMSE, which was less
pronounced in the case of strong quadratic deviations. With increasing 
prediction AUC we saw an increased tendency of the adaptive approach to select
more complex models compared to the base case scenario. In the case of strong 
linear deviations, the linear interaction model was selected the majority of the
time (63\%), while the simpler constant treatment effect model was selected 26\%
of the time (Supplementary Table S7). However, the increased RMSE of the
constant treatment effect model inflated the the RMSE of the linear interaction
model. In the case of strong quadratic deviations the RCS (3 knots) model was
selected 84\% of the time.

@@

The constant effects model, the linear interaction model and models with RCS
smoothing (3 knots) had the highest median c-for-benefit in the base case
scenario and the scenarios where linear and quadratic deviations were
considered. The constant treatment effect model and the linear interaction model
tended to present much lower variability compared to all other approaches
(Figure \ref{fig:discrimination}). With increasing number of RCS knots, we
observed decreasing median values and increasing variability of the
c-for-benefit in all scenarios.

@@

As for calibration, the linear interaction model had the lowest median
ICI-for-benefit in the majority of scenarios except for the scenarios where no
or moderate linear deviations from the base case were considered. In that case
constant treatment effect models demonstrated optimal calibration, very
comparable to the linear interaction model's performance, nonetheless (Figure
\ref{fig:calibration}). With strong linear or quadratic deviations, the
constant treatment effect model was poorly calibrated (Figure
\ref{fig:calibration}; Panels C and D).

>>>>

---
title: |
 Linear interaction of treatment with baseline risk was sufficient for predicting
 individualized benefit
abstract: |
 **Objective:** To compare different risk-based methods predicting individualized
 treatment effects in RCT simulations. **Study Design and Setting:** We simulated
 data using diverse assumptions for a baseline prognostic index of risk (PI) and
 the shape of its interaction with treatment (none, linear or quadratic). In
 each sample we predicted absolute benefit using: models with the PI and a
 constant relative treatment effect; models including an interaction of treatment
 with the PI; stratification in quarters of the PI; nonlinear transformations of
 the PI (restricted cubic splines with 3, 4 and 5 knots); an adaptive approach
 using Akaike’s Information Criterion. We evaluated predictive performance using
 root mean squared error and measures of discrimination and calibration for
 benefit. Starting from a base case scenario (sample size 4,250, treatment odds
 ratio 0.8, AUC of the PI 0.75), we varied the sample size, the treatment effect
 strength, and the Ti's discriminative ability. **Results:** Models including a
 PI by treatment interaction performed better under most simulation settings.
 Flexible models required larger sample sizes and higher AUC of the PI to
 outperform the linear interaction model. The adaptive approach tended to select
 simpler models with smaller sample sizes and/or worse prediction performance.
 **Conclusion:** Under most circumstances, a model with a linear interaction of
 the PI with treatment is the optimal risk-based approach to predicting absolute
 treatment benefit.
authors:
  - name: Alexandros Rekkas
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Peter R. Rijnbeek
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Ewout W. Steyerberg
    department: Department of Biomedical Data Sciences
    affiliation: Leiden University Medical Center
    location: Leiden, The Netherlands
  - name: David van Klaveren
    department: Department of Public Health
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
keywords:
  - treatment effect heterogeneity
  - absolute benefit
  - prediction models
output: 
  rticles::arxiv_article:
    keep_tex: true
  bookdown::word_document2:
    reference_docx: reference.docx
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage[pagewise, left]{lineno}
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{caption}
  - \usepackage{booktabs}
  - \date{}
  - \newcommand\given[1][]{\:#1\vert\:}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

@@

Combining all these settings resulted in a simulation study of 648 scenarios
(exact settings in the supplementary material).

@@

Second, we considered a model which assumes *constant relative treatment effect*
(constant odds ratio). Hence, absolute benefit is predicted from
$\hat{\tau}(\bm{x}) = \text{expit}(PI +\log(\text{OR}))$.

@@

```{r adaptive, echo=FALSE, warning=FALSE, message=FALSE}
adaptiveSelections <- readr::read_csv(here::here("data/processed/adaptiveSelections.csv"))
rmseDistribution <- readr::read_csv(here::here("data/processed/rmseDistribution.csv"))
aucDistribution <- readr::read_csv(here::here("data/processed/discriminationDistribution.csv"))
```
The linear interaction model outperformed all RCS methods in terms of
RMSE in scenarios with true constant relative treatment effect (OR =
0.8, N = `r 4250` and AUC = 0.75), strong linear and even strong
quadratic deviations (Figure \ref{fig:rmsebase}). However, with
non-monotonic deviations errors of the linear interaction model
increased substantially, especially in the presence of
treatment-related harms. In these scenarios, RCS-3 proved to be quite
robust outperforming all other methods. The constant treatment effect
approach had overall best performance under true constant treatment
effect settings, but was sensitive to all considered deviations,
resulting in increased error rates. Finally, the adaptive approach had
comparable performance to the best-performing method in each
scenario. However, we observed increased error variability in the case
of linear and non-monotonic deviations, especially for moderate or
strong treatment-related harms.

@@

Increasing the sample size to `r 17000` favored the RCS-3 approach the
most, as it achieved lowest or close to lowest RMSE across all
scenarios (Figure \ref{fig:rmsesamplesize}). Especially in cases of
strong quadratic and non-monotonic deviations RCS-3 had lower error
rates
(median RMSE
`r rmseDistribution %>% filter(scenarioId == 385) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_rcs_3)`
for strong quadratic deviations and 
`r rmseDistribution %>% filter(scenarioId == 421) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_rcs_3)`
for non-monotonic deviations with no treatment-related harms)
compared to the linear interaction approach
(median RMSE
`r rmseDistribution %>% filter(scenarioId == 385) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_linear)` and 
`r rmseDistribution %>% filter(scenarioId == 421) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_linear)`,
respectively), regardless of treatment-related harms strength. The
issues with the large error variability of the adaptive approach
improved with larger sample sizes.

@@

When we increased the AUC of the true prediction model to 0.85 (OR =
0.8 and N = `r 4250`) the constant effect model outperformed all other
methods in the case of true constant treatment effects, but proved to
be the least robust to deviations. Again, RCS-3 had the lowest
error rates in the case of strong quadratic or monotonic deviations
and very comparable performance to the best-performing linear
interaction model in the case of strong linear deviations
(median RMSE
`r rmseDistribution %>% filter(scenarioId == 297) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_rcs_3)` for RCS-3 compared to 
`r rmseDistribution %>% filter(scenarioId == 297) %>% mutate(across(-one_of("scenarioId"), ~ d(.x, decimals = 3))) %>% pull(median_linear)`
for the linear interaction model). The adaptive approach, though it
performed similar to the best performing method in each scenario, on
average, there was increased variability in error rates in the case of
strong linear and non-monotonic deviations. In these scenarios, the adaptive approach often selected the constant treatment effect model
(`r adaptiveSelections %>% filter(scenarioId == 297) %>% pull(treatment) * 100`\%
and
`r adaptiveSelections %>% filter(scenarioId == 405) %>% pull(treatment) * 100`\%
in the strong linear and non-monotonic deviation scenarios without
treatment-related harms, respectively).

@@

In terms of discrimination for benefit (OR = 0.8, N = `r 4250` and AUC
= 0.75), all methods performed similarly, on average (Figure
\ref{fig:discrimination}). In the case of non-monotonic deviations,
the constant effect model had much lower discriminative performance
compared to the rest of the methods (median AUC of
`r aucDistribution %>% filter(scenarioId == 397) %>% mutate(across(-one_of("scenarioId"), ~ d(.x))) %>% pull(median_constant)` 
for the constant effects model compared to the best-performing RCS-3
with
`r aucDistribution %>% filter(scenarioId == 397) %>% mutate(across(-one_of("scenarioId"), ~ d(.x))) %>% pull(median_rcs_3)`).
The linear interaction model was most stable in compared to the other
methods in terms of error variability. With increasing number of RCS
knots, we observed decreasing median values and increasing variability
of the c-for-benefit in all scenarios.

@@

As for calibration, the constant effects model outperformed all other
models in the case of true constant treatment effects, but it was
miscalibrated for all deviation scenarios (Figure
\ref{fig:calibration}). The linear interaction model showed best or
close to best calibration across all scenarios and only showed worse
calibration compared to RCS-3 in the case of non-monotonic deviations
and treatment-related harms. The adaptive approach was worse
calibrated in scenarios with strong linear and non-monotonic
deviations compared to the linear interaction model and RCS-3.
